Runtime-Assisted Convergence in Replicated Data Types

### 2021-11-19 introduction.tex
Edits to abstract and introduction

% hinders the widespread adoption of replicated data
% types in software development.
In this paper we propose such an alternative approach to convergent
replicated data types.  
% The key distinguishing characteristic of
Our approach differs from the CRDT approach in
two fundamental ways. Firstly, we adopt a state-centric model of
replication with \emph{mergeability} of states as the defining
characteristic in contrast to CRDTs' operation-centric model
characterized by commutativity. Unlike commutativity, mergeability
does not require a full-scale re-engineering of the data type
definition. Instead, a data type only has to define a merge semantics
in form a \C{merge} function to reconcile concurrent versions of the
type in presence of their common ancestor version. Notably, \C{merge}
function is left completely unconstrained even as convergence is
guaranteed on the corresponding \emph{mergeable replicated data type}
(MRDT)~\cite{mrdt}. Such strong guarantees are possible due to the
simplicity of state-based reasoning relative to the operation- or
trace-based one. Note that state-centric model of replication has also
been adapted to CRDTs, but such state-based CRDTs require the
replicated state to be organized as a lattice to guarantee convergence
-- a restriction we eliminate in our setting.

The second distinguishing aspect of our approach is the utilization of a
distributed runtime that orchestrates \emph{well-formed} distributed
executions that are guaranteed to converge. Note that it is easy to
construct a trivial runtime that guarantees convergence of all executions.
Such a runtime would make an extensive use of inter-replica coordination to
synchronize the execution of every operation and thus induce a
sequentially-consistent behavior. This would however defeat the purpose of
state replication as the latency incurred for synchronized execution of an
operation is prohibitively high, and the availability of the system during
the execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to not synchronize the execution of RDT
operations. Orchestrating a well-formed convergent execution without
impacting latency and availability may seem impossible due to the CAP
theorem~\cite{cap}, but it is not the case in fact. While CAP theorem does
rule out linearizability, it does not prohibit enforcing weaker consistency
constraints on distributed executions. In fact many CRDTs, including the
aforementioned \emph{remove-wins} set, implicitly assume \emph{causal
consistency}, which is weaker than linearizability but is stronger than the
default \emph{eventual consistency}. While causal consistency alone cannot
guarantee convergence, it sufficiently constrains the distributed execution
for the CRDT implementations to guarantee convergence. In this respect our
approach differs in the sense that (a). It identifies a novel set of
constraints on distributed execution that guarantee the convergence of an
MRDT \emph{regardless} of its merge semantics, and (b). It localizes those
constraints spatially and temporally into the runtime such that
per-operation latency and system-wide availability remain unaffected.
Sec.~\ref{sec:semantics} explains our runtime in detail.

\paragraph{Contributions} The contributions of this paper are summarized
below:
\begin{itemize}
  \item We present a model of state replication for \emph{mergeable} data
  types, i.e., data types equipped with a \C{merge} function, and identify a
  structural well-formedness condition over distributed executions that is
  sufficient to guarantee their convergence. This is in contrast to
  existing models of replicated data types, where each type is obligated to
  prove its convergence.

  \item We formalize a distributed runtime called \quark that orchestrates
  well-formed distributed executions. We prove the correctness of \quark,
  and also demonstrate its \emph{progress} property, namely that no
  distributed execution would ever get ``stuck''. Besides the manual proofs
  for the theorems, we also provide the full formalization of the \quark
  system and its properties in Alloy~\cite{alloy} and Ivy~\cite{ivy}. The
  Ivy formalization of \quark has been automatically verified with help of
  Z3~\cite{z3}. To the best of our knowledge, this is the first time an SMT
  solver was used to verify the \emph{meta-theory} of a formal setup.

  \item We realize an implementation of \quark as a light-weight shim layer
  on top of Scylla -- an off-the-shelf distributed key-value
  store~\cite{scylla}. We do a thorough evaluation with a
    collaborative-editing benchmark to demonstrate that \quark has
    \emph{no} effect on the latency and availability of the system, while
    slightly increasing the time to convergence.
\end{itemize}

>>>>

% hinders the widespread adoption of replicated data
% types in software development.
In this paper we propose such an alternative approach to convergent
replicated data types. The key distinguishing characteristic of our
approach is the utilization of a distributed runtime to orchestrate
\emph{well-formed} distributed executions that are guaranteed to
converge. We define well-formedness as a constraint over the
\emph{states} that arise during the execution as opposed to the
\emph{operations} that are involved in the execution. To allow for its
enforcement, we adopt a state-centric model of replication based on
version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
operation-centric model based on commutative operations. Unlike
commutativity, mergeability does not require a data type definition to
be refactored to suit distributed execution. Instead, the type
definitions needs to be \emph{extended} with a \C{merge} function to
reconcile concurrent versions of the type in presence of their common
ancestor version. Notably, \C{merge} function is left completely
unconstrained. Consequently, programmer can chose any merge semantics
that make sense within the context of the data type's (sequential)
semantics without risking the loss of convergence. Such strong
guarantees are made possible by the simplicity of state-based
reasoning combined with the run-time enforcement of well-formedness.

% combined with the simplicity
% of state-based reasoning. Note that state-centric model of
% replication has also been explored in the context of CRDTs. However,
% state-based CRDTs require the replicated state to be organized as a
% lattice -- a restriction we relax in our setting.

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions.  Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to not synchronize the execution of RDT
operations. Orchestrating a well-formed convergent execution without
impacting latency and availability may seem impossible due to the CAP
theorem~\cite{cap}, but it is not the case in fact. While CAP theorem
does rule out linearizability, it does not prohibit enforcing weaker
consistency constraints on distributed executions. Our approach
exploits this observation by (\rom{1}). Identifying a novel set of
constraints on distributed executions that guarantee their
convergence, and (\rom{2}). Localizing such constraints on
(asynchronously executed) state merges such that per-operation latency
and system-wide availability remain unaffected.
Sec.~\ref{sec:concrete-sem} explains our runtime in detail.

% In fact many CRDTs,
% including the aforementioned \emph{remove-wins} set, implicitly assume
% \emph{causal consistency}, which is weaker than linearizability but is
% stronger than the default \emph{eventual consistency}. While causal
% consistency alone cannot guarantee convergence, it sufficiently
% constrains the distributed execution for the CRDT implementations to
% guarantee convergence. In this respect our approach differs in the
% sense that

\paragraph{Contributions} The contributions of this paper are summarized
below:
\begin{itemize}
  \item We present a runtime-assisted approach to state replication
    that sufficiently constrains distributed executions so as to
    guarantee the convergence of replicated data types. This is in
    contrast to the existing models of replication, where each type is
    obligated to prove its convergence.

  \item We formalize the aforementioned constraints in the context of
    a an abstract machine (named \quark) that manifests distributed
    executions as version history graphs (similar to Git~\cite{git}).
    We prove the \emph{convergence} and \emph{progress} properties of
    the \quark abstract machine that together guarantee its soundness.
    The formalization has been mechanized in Ivy~\cite{ivy} and
    \emph{automatically} verified with help of Z3~\cite{z3}. To the
    best of our knowledge, this is the first time an SMT solver was
    used to verify the \emph{meta-theory} of a formal system.

  \item We formalize \quark distributed machine that implements the
    semantics of \quark abstract machine in the context of an
    asynchronous distributed system. The formalization addresses
    several practical concerns of distribution and serves as a
    blueprint for an efficient implementation.

  \item We describe an implementation of \quark as a light-weight shim
    layer on top of Scylla -- an off-the-shelf distributed key-value
    store~\cite{scylla}. We evaluate the implementation on a
    collaborative-editing benchmark to demonstrate novel tradeoffs
    presented by our approach.
\end{itemize}

### 2021-11-19 quark.tex
Introduction added
Motivation in progress
Motivation text complete
Sec 3.
Quark concrete semantics added
Concrete semantics (Sec 4.)
Edits to abstract and introduction

%% Title information
\title[Convergence Sans Commutativity]{Convergence Sans Commutativity:
Mergeability as a basis for Convergent Replicated Data Types}      %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
   Programming geo-distributed replicated systems is hard given the
   complexity of reasoning about constantly evolving states at different
   replicas. Uncoordinated updates to replicas lead to states that are
   unlikely to converge. In the absence of inter-replica coordination,
   commutativity has come to be accepted as the \emph{de facto} design
   principle to guarantee convergence. Unfortunately, designing replicated
   state updates as commutative operations is hard considering that most
   data types define operations that do not commute. Deriving a replicated
   variant of a data type often entails a creative re-engineering of its
   internal representation and algorithms so as to support commutative
   operations while retaining (close to) the original semantics. In this
   paper we propose an alternative approach to replicated data types that
   avoids such complex re-engineering. Our approach is fundamentally
   different in that it is based on the weaker notion of
   \emph{mergeability}. Unlike commutativity, mergeability only requires a
   data type to define \emph{a} merge semantics while leaving its existing
   definition intact. Notably, our approach leaves type and merge semantics
   completely unconstrained even as it guarantees eventual convergence and
   full availability in an asynchronous distributed setting. In other
   words, there is no need for the programmers to prove algebraic
   properties of their implementations in order to obtain the convergence
   and availability guarantees under a distributed execution. Such
   weakening of obligations is made possible by a novel distributed runtime
   that orchestrates distributed executions as per a structural
   well-formedness criterion, which we show is sufficient to guarantee
   convergence. We implement the runtime called \quark as a thin shim layer
   on top of an off-the-shelf distributed database, and use it to run
   mergeable replicated variants of several ordinary data types. Our
   evaluation categorically demonstrate the generality of our approach and
   the low overhead imposed by our runtime.

\end{abstract}

@@

Text of paper \ldots

>>>>

%% Title information
\title[]{Runtime-Assisted Convergence in Replicated Data Types}
%% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}

  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data
  type operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.
  The cornerstone of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of per-operation latency and system-wide
  availability. Instead, the tradeoff is the time required to achieve
  convergence -- a metric we call \emph{staleness}. We quantify this
  tradeoff via a thorough evaluation of a collaborative editing
  application. and demonstrate a novel practical path towards
  achieving high scalability through geo-replication.
\end{abstract}

@@

\input{motivation}

\input{vc-semantics}

\input{concrete-semantics}
